---
title: "PS270 Final Project"
author: "Joseph Kutrieb"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction
  My research question is how does someone's education level impact their support for either Trump or Kamala in the 2024 presidential election? My hypothesis is that those who have a higher education level will tend to support Kamala more than they will support Trump. This can be important because if this poll were to show that the vast majority of people support Kamala over Trump but also the majority of the people who took this poll have a high level of education, the results showing significant support for Kamala might not be as credible.
  
## Data
  The data I will be using is an election opinion poll from Pew Research Center from just prior to the 2024 election gauging support for Trump and Kamala as well as various issues like healthcare, economy, and the Supreme Court. First I will look at the proportion of each education level of the participants in this particular poll as well as the proportion of support for Trump and Kamala indpendent of education level. Then I will sort this between the proportion of support for Kamala and the proportion of support for Trump. 
  
```{r, echo = FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
data <- read_csv("ATP W153.csv")
```


```{r, echo = FALSE}
data <- data |>
  mutate(education_level = case_when(
    F_EDUCCAT2 == 1 ~ "Less than HS",
    F_EDUCCAT2 == 2 ~ "HS Grad",
    F_EDUCCAT2 == 3 ~ "Some College",
    F_EDUCCAT2 == 4 ~ "Assoc Degree",
    F_EDUCCAT2 == 5 ~ "College Grad",
    F_EDUCCAT2 == 6 ~ "Postgrad",
    F_EDUCCAT2 == 99 ~ "Refused"
  )) |>
  mutate(education_level = factor(education_level, 
                             levels = c("Less than HS", "HS Grad", "Some College", "Assoc Degree", "College Grad", "Postgrad", "Refused")))

data <- data |>
  mutate(candidate_support = case_when(
    VOTEGEN24_W153 == 1 ~ "Trump",
    VOTEGEN24_W153 == 2 ~ "Kamala",
    VOTEGEN24_W153 == 4 ~ "None/Other",
    VOTEGEN24_W153 == 99 ~ "Refused"
  )) |>
  mutate(candidate_support = factor(candidate_support, levels = c("Trump", "Kamala", "None/Other", "Refused")))


education <-
  data |> 
  group_by(education_level) |>
  summarize(n = n()) |>
  mutate(prop = n / sum(n))
  
knitr::kable(education, digits = 2, col.names = c("Education Level", "Number of Respondents", "Proportion"))

candidate  <-
  data |>
  filter(candidate_support == c("Kamala", "Trump", "None/Other", "Refused")) |>
  group_by(candidate_support) |>
  summarize(n = n()) |>
  mutate(prop = n/ sum(n))

knitr::kable(candidate, digits = 2, col.names = c("Candidate Support", "Number of Respondents", "Proportion"))




```




```{r, echo = FALSE}
education_barplot <- 
  ggplot(education, aes(x = education_level, prop)) + 
  geom_col() +
  labs(x = "Level of Education", y = "Proportion", title = "Proportion of Varying Levels of Education in Election Polling", caption = "Source: Pew Research Center")

education_barplot

candidate_barplot <- 
  ggplot(candidate, aes(x = candidate_support, prop)) + 
  geom_col() +
  labs(x = "Support for Trump and Kamala", y = "Proportion", title = "Proportion of Support for Trump and Kamala", caption = "Source: Pew Research Center")

candidate_barplot

```
These tables and plots show the overall proportion of each variable independent of each other. Based on these alone, it can be seen that the majority of respondents support Kamala and that there is a significant number who have received some sort of college education. This alone is not enough to prove anything and an analysis of the proportion of each education level and who they support is needed.

# Results
This table and barplot both show the proportion of support for either Trump or Kamala based on education level, and there seems to be a trend of as people's highest level of education increases, their support for Kamala also increases.
  
```{r, echo = FALSE}
educan_table <- 
  data |>
  filter(candidate_support == c("Trump", "Kamala")) |>
  filter(education_level == c("Less than HS", "HS Grad", "Some College", "Assoc Degree", "College Grad", "Postgrad")) |>
  group_by(education_level, candidate_support) |>
  summarize(n = n()) |>
  mutate(prop = n / sum(n)) |>
  select(-n)

knitr::kable(educan_table, digits = 2, col.names = c("Education Level", "Candidate Support", "Proportion"))


ggplot(educan_table, aes(x = education_level, y = prop, fill = candidate_support)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(Kamala = "steelblue1", Trump = "indianred1")) +
  labs(x = "Level of Education", y = "Proportion", title = "Proportion of Support for Trump and Kamala Based on Education", caption = "Source: Pew Research Center", fill = "Candidates")
```
This alone is again not enough to definitively prove the hypothesis as it at most shows a trend but that is not enough to answer the question being asked. For this a regression model is needed, and a linear regression will be the best fit for this.

```{r, echo = FALSE}
library(modelsummary)

lm_model <-
  lm(candidate_support ~ education_level, data = data, family = binomial)


modelsummary(lm_model, coef_rename = c("education_levelHS Grad" = "HS Grad", "education_levelSome College" = "Some College", "education_levelAssoc Degree" = "Assoc Degree", "education_levelCollege Grad" = "College Grad", "education_levelPostgrad" = "Postgrad", "education_levelRefused" = "Refused"))

```
This regression output gives further evidence of the relationship between a higher level of education and support for Kamala. In the data, support for Trump is listed as 1 while support for Kamala is listed as 2 so an intercept of 1.601 shows more support for Kamala than for Trump. The coefficients show the percent that lean towards either Trump or Kamala. For example, those with a post graduate education are 16.7% more likely to support Kamala while those with only a high school graduation are 2.8% more likely to support Trump. To prove this hypothesis though, it is necessary to reject the null hypothesis. For this I will be listing anyone with an associates degree or higher as having "Higher Education" and those with only some college or a high school degree or neither as "No Higher Education".

```{r, echo = FALSE}
library(infer)

data_clean <- data |>
  mutate(
    candidate_support = as.character(candidate_support),
    education_2 = case_when(
      F_EDUCCAT2 == 1:3 ~ "No Higher Education",
      F_EDUCCAT2 == 4:6 ~ "Higher Education",
      F_EDUCCAT2 == 99 ~ "Refused")) |>
  filter(candidate_support == c("Trump", "Kamala"),
         education_2 == c("No Higher Education", "Higher Education"))


data <-
  data |>
  mutate(education_2 = case_when(
    F_EDUCCAT2 == 1 ~ "No Higher Education",
    F_EDUCCAT2 == 2 ~ "No Higher Education",
    F_EDUCCAT2 == 3 ~ "No Higher Education",
    F_EDUCCAT2 == 4 ~ "Higher Education",
    F_EDUCCAT2 == 5 ~ "Higher Education",
    F_EDUCCAT2 == 6 ~ "Higher Education",
    F_EDUCCAT2 == 99 ~ "Refused"))

edu_diff <-
  data_clean |>
  mutate(education_2 = case_when(
    F_EDUCCAT2 == 1 ~ "No Higher Education",
    F_EDUCCAT2 == 2 ~ "No Higher Education",
    F_EDUCCAT2 == 3 ~ "No Higher Education",
    F_EDUCCAT2 == 4 ~ "Higher Education",
    F_EDUCCAT2 == 5 ~ "Higher Education",
    F_EDUCCAT2 == 6 ~ "Higher Education",
    F_EDUCCAT2 == 99 ~ "Refused")) |>
  filter(candidate_support %in% c("Trump", "Kamala"), education_2 %in% c("No Higher Education", "Higher Education")) |>
  specify(candidate_support ~ education_2, success = "Kamala") |>
  calculate(stat = "diff in props", order = c("Higher Education", "No Higher Education"))


null_dist <-
  data_clean |>
  mutate(education_2 = case_when(
    F_EDUCCAT2 == 1 ~ "No Higher Education",
    F_EDUCCAT2 == 2 ~ "No Higher Education",
    F_EDUCCAT2 == 3 ~ "No Higher Education",
    F_EDUCCAT2 == 4 ~ "Higher Education",
    F_EDUCCAT2 == 5 ~ "Higher Education",
    F_EDUCCAT2 == 6 ~ "Higher Education",
    F_EDUCCAT2 == 99 ~ "Refused")) |>
  filter(candidate_support %in% c("Trump", "Kamala"), education_2 %in% c("No Higher Education", "Higher Education")) |>
  specify(candidate_support ~ education_2, success = "Kamala") |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in props", order = c("Higher Education", "No Higher Education"))

p_val <-
  null_dist |>
  get_p_value(obs_stat = edu_diff, direction = "two-sided")

knitr::kable(p_val, col.names = c("P-Value"))


```
The p-value here is 0 and with an alpha value of 0.05 this shows that the null hypothesis can easily be rejected. The result of the effect of education on preference for either Trump or Kamala is very clear here.

## Conclusion
Based on this data, it can safely be said that a higher level of education leads to someone having a greater preference for Kamala Harris over Donald Trump. The plot showing the proportions, the linear regression, and the p-value all show that there education has a significant effect. This analysis does not account for confounding factors that could also be influencing support for one candidate or another. Some confounding factors that could be accounted for if there was more information could be geographic location, wealth, and age which could all influence preference for one candidate or another. Had I more time, money, and access to more data such as where the respondents live, certain confounding factors and bias's could be accounted for and ruled out.
  

